---
title: "The Ramble Register — exploratory analysis"
subtitle: "Acoustic bird detections from the Upper West Side"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110
)
```

This notebook connects to the same PostgreSQL database the listener writes to and
the dashboard reads from, pulls every detection into a data frame, and explores
it. It reuses the project's `.env` for credentials, so **nothing secret lives in
this file** — it is safe to commit.

> Read-only by design: every query here is a `SELECT`. Point it at a read-only
> DB role if you have one.

## 1. Install the R packages (run once)

This chunk does not run automatically. Run it by hand the first time, then leave
it be.

```{r install, eval=FALSE}
install.packages(c(
  "DBI", "RPostgres",   # database connection
  "dplyr", "tidyr",     # wrangling
  "lubridate",          # timestamps
  "ggplot2", "scales",  # plots
  "knitr"               # tables
))
```

```{r libraries}
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(scales)
library(knitr)
```

## 2. Credentials & connection

The project keeps its Postgres credentials in a `.env` file next to this
notebook (the same one `config.py` loads). We read it here so they never have to
be typed into — or saved in — this document. `.env` is git-ignored.

If you ran the listener/dashboard locally, `.env` already exists. Otherwise copy
`.env.example` to `.env` and fill in `DB_HOST`, `DB_NAME`, `DB_USER`, `DB_PASS`
(and `DB_PORT` / `BW_DB_SCHEMA` if non-default).

```{r read-env}
# Load .env (KEY=VALUE lines) into the R session environment, just like
# python-dotenv does for the rest of the project.
if (file.exists(".env")) {
  readRenviron(".env")
} else {
  warning("No .env found — relying on variables already set in the environment.")
}

# Pull the same variables config.py uses, with the same defaults.
db_host   <- Sys.getenv("DB_HOST",      "localhost")
db_port   <- Sys.getenv("DB_PORT",      "5432")
db_name   <- Sys.getenv("DB_NAME",      "birdwatcher")
db_user   <- Sys.getenv("DB_USER",      "")
db_pass   <- Sys.getenv("DB_PASS",      "")
db_schema <- Sys.getenv("BW_DB_SCHEMA", "birdwatcher")
tz        <- Sys.getenv("BW_TIMEZONE",  "America/New_York")

# If a full SQLAlchemy-style URL was supplied instead of the parts, unpack it.
# Form: postgresql+psycopg2://user:pass@host:port/dbname
db_url <- Sys.getenv("BW_DATABASE_URL", "")
if (nzchar(db_url)) {
  m <- regmatches(
    db_url,
    regexec("://([^:@/]+)(?::([^@/]*))?@([^:/]+)(?::(\\d+))?/(.+)$", db_url)
  )[[1]]
  if (length(m) == 6) {
    db_user <- m[2]; db_pass <- m[3]; db_host <- m[4]
    db_port <- ifelse(nzchar(m[5]), m[5], db_port); db_name <- m[6]
  }
}

stopifnot(nzchar(db_user))  # fail early with a clear message if .env is empty
```

```{r connect}
con <- dbConnect(
  RPostgres::Postgres(),
  host     = db_host,
  port     = as.integer(db_port),
  dbname   = db_name,
  user     = db_user,
  password = db_pass
)

# Scope unqualified table names to the birdwatcher schema (validated to a safe
# identifier first, mirroring db.py's _ident check).
stopifnot(grepl("^[A-Za-z0-9_]+$", db_schema))
dbExecute(con, sprintf('SET search_path TO "%s"', db_schema))
```

## 3. Pull the data

Detections are stored as timezone-naive **local** wall-clock time (see
`config.now_local`). We cast to text in SQL and re-parse with the project's
timezone so hour-of-day buckets are exactly right regardless of the R session's
locale.

```{r load-detections}
det <- dbGetQuery(con, "
  SELECT id,
         detected_at::text AS detected_at,
         common_name,
         scientific_name,
         confidence,
         clip_path,
         clip_url
  FROM detections
  ORDER BY detected_at
")

det <- det %>%
  mutate(
    detected_at = ymd_hms(detected_at, tz = tz),
    date        = as_date(detected_at),
    hour        = hour(detected_at),
    weekday     = wday(detected_at, label = TRUE, week_start = 1),
    has_clip    = !is.na(clip_url)
  )

nrow(det)
```

```{r load-votes}
# Clip up/down votes (may be empty if no one has voted yet).
votes <- dbGetQuery(con, "
  SELECT detection_id, voter, vote, created_at::text AS created_at
  FROM clip_votes
")
nrow(votes)
```

## 4. Overview

```{r overview}
overview <- tibble(
  metric = c("Total detections", "Distinct species", "Days with activity",
             "First detection", "Latest detection", "Clips with audio"),
  value  = c(
    format(nrow(det), big.mark = ","),
    as.character(n_distinct(det$common_name)),
    as.character(n_distinct(det$date)),
    as.character(min(det$detected_at)),
    as.character(max(det$detected_at)),
    sprintf("%d (%.0f%%)", sum(det$has_clip), 100 * mean(det$has_clip))
  )
)
kable(overview, col.names = c("", ""))
```

## 5. Most-heard species

```{r top-species}
top_species <- det %>%
  count(common_name, name = "detections") %>%
  arrange(desc(detections)) %>%
  slice_head(n = 20)

ggplot(top_species, aes(reorder(common_name, detections), detections)) +
  geom_col(fill = "#2e7d5b") +
  coord_flip() +
  labs(title = "Top 20 species by number of detections",
       x = NULL, y = "Detections") +
  theme_minimal(base_size = 12)
```

## 6. Activity over time

```{r daily-trend}
daily <- det %>% count(date, name = "detections")

ggplot(daily, aes(date, detections)) +
  geom_col(fill = "#2e7d5b") +
  scale_x_date(labels = label_date_short()) +
  labs(title = "Detections per day", x = NULL, y = "Detections") +
  theme_minimal(base_size = 12)
```

```{r cumulative-species}
# When was each species first heard? A cumulative "life list" curve.
first_seen <- det %>%
  group_by(common_name) %>%
  summarise(first = min(date), .groups = "drop") %>%
  arrange(first) %>%
  mutate(cumulative_species = row_number())

ggplot(first_seen, aes(first, cumulative_species)) +
  geom_step(color = "#2e7d5b", linewidth = 0.8) +
  labs(title = "Cumulative species heard (life list growth)",
       x = NULL, y = "Distinct species") +
  theme_minimal(base_size = 12)
```

## 7. Daily rhythm (when are birds active?)

```{r diurnal}
hourly <- det %>% count(hour, name = "detections")

ggplot(hourly, aes(hour, detections)) +
  geom_col(fill = "#2e7d5b") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Detections by hour of day (all dates pooled)",
       x = "Hour", y = "Detections") +
  theme_minimal(base_size = 12)
```

```{r hour-weekday-heatmap}
# Hour-of-day × day-of-week heatmap — surfaces dawn-chorus and weekend effects.
heat <- det %>% count(weekday, hour, name = "detections")

ggplot(heat, aes(hour, weekday, fill = detections)) +
  geom_tile() +
  scale_fill_gradient(low = "#eef6f1", high = "#1b5e3f") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Activity by hour and weekday", x = "Hour", y = NULL,
       fill = "Detections") +
  theme_minimal(base_size = 12)
```

## 8. Confidence

```{r confidence-dist}
ggplot(det, aes(confidence)) +
  geom_histogram(binwidth = 0.02, fill = "#2e7d5b") +
  labs(title = "Detection confidence distribution",
       x = "BirdNET confidence", y = "Detections") +
  theme_minimal(base_size = 12)
```

```{r confidence-by-species}
# Median confidence for the most-heard species — low medians are worth a
# listen/vote before trusting the ID.
det %>%
  group_by(common_name) %>%
  summarise(detections = n(),
            median_conf = median(confidence),
            best_conf   = max(confidence),
            .groups = "drop") %>%
  filter(detections >= 10) %>%
  arrange(median_conf) %>%
  slice_head(n = 15) %>%
  kable(digits = 3,
        col.names = c("Species", "Detections", "Median conf.", "Best conf."))
```

## 9. The life list

```{r life-list}
det %>%
  group_by(common_name, scientific_name) %>%
  summarise(detections = n(),
            first_seen  = min(detected_at),
            last_seen   = max(detected_at),
            best_conf   = max(confidence),
            .groups = "drop") %>%
  arrange(first_seen) %>%
  kable(digits = 3,
        col.names = c("Species", "Scientific name", "Detections",
                      "First heard", "Last heard", "Best conf."))
```

## 10. Crowd verdicts on clips

```{r voting, eval=TRUE}
if (nrow(votes) > 0) {
  tally <- votes %>%
    group_by(detection_id) %>%
    summarise(up   = sum(vote > 0),
              down = sum(vote < 0),
              net  = sum(vote), .groups = "drop")

  voted <- det %>% inner_join(tally, by = c("id" = "detection_id"))

  cat(sprintf("%d clips have at least one vote.\n", nrow(voted)))

  voted %>%
    arrange(net) %>%
    transmute(common_name, detected_at, confidence, up, down, net) %>%
    slice_head(n = 15) %>%
    kable(digits = 3,
          caption = "Most-disputed clips (lowest net score first)")
} else {
  cat("No clip votes recorded yet.")
}
```

## 11. Disconnect

```{r disconnect}
dbDisconnect(con)
```

---

*The Ramble Register — acoustic bird monitoring, Upper West Side, Manhattan.*
*Edit the queries above or add your own chunks; `det` and `votes` hold all the data.*
