---
title: "The Ramble Register — exploratory analysis"
subtitle: "Acoustic bird detections from the Upper West Side"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110
)
```

This notebook connects to the same PostgreSQL database the listener writes to and
the dashboard reads from, pulls every detection into a data frame, and explores
it. It reuses the project's `.env` for credentials, so **nothing secret lives in
this file** — it is safe to commit.

> Read-only by design: every query here is a `SELECT`. Point it at a read-only
> DB role if you have one.

## 1. Install the R packages (run once)

This chunk does not run automatically. Run it by hand the first time, then leave
it be.

```{r install, eval=FALSE}
install.packages(c(
  "DBI", "RPostgres",   # database connection
  "dplyr", "tidyr",     # wrangling
  "lubridate",          # timestamps
  "ggplot2", "scales",  # plots
  "knitr",              # tables
  "isotree"             # optional: isolation-forest outlier detection (§12)
))
```

```{r libraries}
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(scales)
library(knitr)
```

## 2. Credentials & connection

The project keeps its Postgres credentials in a `.env` file next to this
notebook (the same one `config.py` loads). We read it here so they never have to
be typed into — or saved in — this document. `.env` is git-ignored.

If you ran the listener/dashboard locally, `.env` already exists. Otherwise copy
`.env.example` to `.env` and fill in `DB_HOST`, `DB_NAME`, `DB_USER`, `DB_PASS`
(and `DB_PORT` / `BW_DB_SCHEMA` if non-default).

```{r read-env}
# Load .env (KEY=VALUE lines) into the R session environment, just like
# python-dotenv does for the rest of the project.
if (file.exists(".env")) {
  readRenviron(".env")
} else {
  warning("No .env found — relying on variables already set in the environment.")
}

# Pull the same variables config.py uses, with the same defaults.
db_host   <- Sys.getenv("DB_HOST",      "localhost")
db_port   <- Sys.getenv("DB_PORT",      "5432")
db_name   <- Sys.getenv("DB_NAME",      "birdwatcher")
db_user   <- Sys.getenv("DB_USER",      "")
db_pass   <- Sys.getenv("DB_PASS",      "")
db_schema <- Sys.getenv("BW_DB_SCHEMA", "birdwatcher")
tz        <- Sys.getenv("BW_TIMEZONE",  "America/New_York")

# If a full SQLAlchemy-style URL was supplied instead of the parts, unpack it.
# Form: postgresql+psycopg2://user:pass@host:port/dbname
db_url <- Sys.getenv("BW_DATABASE_URL", "")
if (nzchar(db_url)) {
  m <- regmatches(
    db_url,
    regexec("://([^:@/]+)(?::([^@/]*))?@([^:/]+)(?::(\\d+))?/(.+)$", db_url)
  )[[1]]
  if (length(m) == 6) {
    db_user <- m[2]; db_pass <- m[3]; db_host <- m[4]
    db_port <- ifelse(nzchar(m[5]), m[5], db_port); db_name <- m[6]
  }
}

stopifnot(nzchar(db_user))  # fail early with a clear message if .env is empty
```

```{r connect}
con <- dbConnect(
  RPostgres::Postgres(),
  host     = db_host,
  port     = as.integer(db_port),
  dbname   = db_name,
  user     = db_user,
  password = db_pass
)

# Scope unqualified table names to the birdwatcher schema (validated to a safe
# identifier first, mirroring db.py's _ident check).
stopifnot(grepl("^[A-Za-z0-9_]+$", db_schema))
dbExecute(con, sprintf('SET search_path TO "%s"', db_schema))
```

## 3. Pull the data

Detections are stored as timezone-naive **local** wall-clock time (see
`config.now_local`). We cast to text in SQL and re-parse with the project's
timezone so hour-of-day buckets are exactly right regardless of the R session's
locale.

```{r load-detections}
det <- dbGetQuery(con, "
  SELECT id,
         detected_at::text AS detected_at,
         common_name,
         scientific_name,
         confidence,
         clip_path,
         clip_url
  FROM detections
  ORDER BY detected_at
")

det <- det %>%
  mutate(
    detected_at = ymd_hms(detected_at, tz = tz),
    date        = as_date(detected_at),
    hour        = hour(detected_at),
    weekday     = wday(detected_at, label = TRUE, week_start = 1),
    has_clip    = !is.na(clip_url)
  )

nrow(det)
```

```{r load-votes}
# Clip up/down votes (may be empty if no one has voted yet).
votes <- dbGetQuery(con, "
  SELECT detection_id, voter, vote, created_at::text AS created_at
  FROM clip_votes
")
nrow(votes)
```

## 4. Overview

```{r overview}
overview <- tibble(
  metric = c("Total detections", "Distinct species", "Days with activity",
             "First detection", "Latest detection", "Clips with audio"),
  value  = c(
    format(nrow(det), big.mark = ","),
    as.character(n_distinct(det$common_name)),
    as.character(n_distinct(det$date)),
    as.character(min(det$detected_at)),
    as.character(max(det$detected_at)),
    sprintf("%d (%.0f%%)", sum(det$has_clip), 100 * mean(det$has_clip))
  )
)
kable(overview, col.names = c("", ""))
```

## 5. Most-heard species

```{r top-species}
top_species <- det %>%
  count(common_name, name = "detections") %>%
  arrange(desc(detections)) %>%
  slice_head(n = 20)

ggplot(top_species, aes(reorder(common_name, detections), detections)) +
  geom_col(fill = "#2e7d5b") +
  coord_flip() +
  labs(title = "Top 20 species by number of detections",
       x = NULL, y = "Detections") +
  theme_minimal(base_size = 12)
```

## 6. Activity over time

```{r daily-trend}
daily <- det %>% count(date, name = "detections")

ggplot(daily, aes(date, detections)) +
  geom_col(fill = "#2e7d5b") +
  scale_x_date(labels = label_date_short()) +
  labs(title = "Detections per day", x = NULL, y = "Detections") +
  theme_minimal(base_size = 12)
```

```{r cumulative-species}
# When was each species first heard? A cumulative "life list" curve.
first_seen <- det %>%
  group_by(common_name) %>%
  summarise(first = min(date), .groups = "drop") %>%
  arrange(first) %>%
  mutate(cumulative_species = row_number())

ggplot(first_seen, aes(first, cumulative_species)) +
  geom_step(color = "#2e7d5b", linewidth = 0.8) +
  labs(title = "Cumulative species heard (life list growth)",
       x = NULL, y = "Distinct species") +
  theme_minimal(base_size = 12)
```

## 7. Daily rhythm (when are birds active?)

```{r diurnal}
hourly <- det %>% count(hour, name = "detections")

ggplot(hourly, aes(hour, detections)) +
  geom_col(fill = "#2e7d5b") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Detections by hour of day (all dates pooled)",
       x = "Hour", y = "Detections") +
  theme_minimal(base_size = 12)
```

```{r hour-weekday-heatmap}
# Hour-of-day × day-of-week heatmap — surfaces dawn-chorus and weekend effects.
heat <- det %>% count(weekday, hour, name = "detections")

ggplot(heat, aes(hour, weekday, fill = detections)) +
  geom_tile() +
  scale_fill_gradient(low = "#eef6f1", high = "#1b5e3f") +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Activity by hour and weekday", x = "Hour", y = NULL,
       fill = "Detections") +
  theme_minimal(base_size = 12)
```

## 8. Confidence

```{r confidence-dist}
ggplot(det, aes(confidence)) +
  geom_histogram(binwidth = 0.02, fill = "#2e7d5b") +
  labs(title = "Detection confidence distribution",
       x = "BirdNET confidence", y = "Detections") +
  theme_minimal(base_size = 12)
```

```{r confidence-by-species}
# Median confidence for the most-heard species — low medians are worth a
# listen/vote before trusting the ID.
det %>%
  group_by(common_name) %>%
  summarise(detections = n(),
            median_conf = median(confidence),
            best_conf   = max(confidence),
            .groups = "drop") %>%
  filter(detections >= 10) %>%
  arrange(median_conf) %>%
  slice_head(n = 15) %>%
  kable(digits = 3,
        col.names = c("Species", "Detections", "Median conf.", "Best conf."))
```

## 9. The life list

```{r life-list}
det %>%
  group_by(common_name, scientific_name) %>%
  summarise(detections = n(),
            first_seen  = min(detected_at),
            last_seen   = max(detected_at),
            best_conf   = max(confidence),
            .groups = "drop") %>%
  arrange(first_seen) %>%
  kable(digits = 3,
        col.names = c("Species", "Scientific name", "Detections",
                      "First heard", "Last heard", "Best conf."))
```

## 10. Crowd verdicts on clips

```{r voting, eval=TRUE}
if (nrow(votes) > 0) {
  tally <- votes %>%
    group_by(detection_id) %>%
    summarise(up   = sum(vote > 0),
              down = sum(vote < 0),
              net  = sum(vote), .groups = "drop")

  voted <- det %>% inner_join(tally, by = c("id" = "detection_id"))

  cat(sprintf("%d clips have at least one vote.\n", nrow(voted)))

  voted %>%
    arrange(net) %>%
    transmute(common_name, detected_at, confidence, up, down, net) %>%
    slice_head(n = 15) %>%
    kable(digits = 3,
          caption = "Most-disputed clips (lowest net score first)")
} else {
  cat("No clip votes recorded yet.")
}
```

## 11. Hypothesis: is some "birdsong" actually construction noise?

Construction noise is constrained in ways real birds are not. In NYC the default
permitted window is **weekday daytime, 7am–6pm** — so construction is essentially
absent at **dawn** and on **weekends**. Songbirds are the opposite: they peak in
the dawn chorus and sing every day of the week.

So a species whose detections cluster in weekday business hours, while staying
nearly silent at dawn and on weekends, behaves like a machine, not a bird. We
flag those.

```{r construction-window}
# NYC's default construction window. Adjust if you know the local pattern.
biz_start <- 7L   # 7am
biz_end   <- 18L  # 6pm (exclusive)

det <- det %>%
  mutate(
    is_weekend  = wday(detected_at, week_start = 1) >= 6,          # Sat/Sun
    is_business = !is_weekend & hour >= biz_start & hour < biz_end,
    is_dawn     = hour >= 4 & hour < 7                             # pre-work dawn
  )

# If detections were spread evenly across the whole week, what fraction would
# land in the business-hours window? This is the null we test each species against.
baseline_business <- ((biz_end - biz_start) * 5) / (24 * 7)
sprintf("Even-spread baseline for the business-hours window: %.1f%%",
        100 * baseline_business)
```

```{r construction-suspects}
suspect <- det %>%
  group_by(common_name) %>%
  summarise(
    n          = n(),
    n_business = sum(is_business),
    weekend_share  = mean(is_weekend),
    dawn_share     = mean(is_dawn),
    median_conf    = median(confidence),
    .groups = "drop"
  ) %>%
  mutate(business_share = n_business / n) %>%
  filter(n >= 20) %>%   # need enough detections to say anything
  rowwise() %>%
  mutate(
    # One-sided binomial test: is this species' business-hours share *higher*
    # than the even-spread baseline by more than chance?
    p_business = binom.test(n_business, n, p = baseline_business,
                            alternative = "greater")$p.value
  ) %>%
  ungroup() %>%
  mutate(
    # "Behaves like construction": daytime-heavy, weekend-light, dawn-light, and
    # the daytime skew is statistically clear.
    construction_suspect = business_share > 0.60 &
                           weekend_share  < 0.10 &
                           dawn_share     < 0.05 &
                           p_business     < 0.01
  ) %>%
  arrange(desc(business_share))

suspect %>%
  transmute(common_name, n,
            business_share, weekend_share, dawn_share,
            median_conf, p_business, construction_suspect) %>%
  kable(digits = 3,
        col.names = c("Species", "n", "Business hrs", "Weekend", "Dawn",
                      "Median conf.", "p (vs baseline)", "Suspect?"),
        caption = sprintf(
          "Per-species timing. Suspect = >60%% in business hours, <10%% weekend, <5%% dawn, p<0.01 (baseline %.0f%%).",
          100 * baseline_business))
```

```{r construction-scatter}
# The tell-tale corner: high business-hours share AND low weekend share
# (bottom-right) is where construction-like sources sit. Real birds cluster
# up and to the left.
ggplot(suspect, aes(business_share, weekend_share)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0.60, linetype = "dashed", color = "grey60") +
  geom_point(aes(size = n, color = construction_suspect), alpha = 0.8) +
  geom_text(data = ~filter(.x, construction_suspect),
            aes(label = common_name), vjust = -0.9, size = 3,
            check_overlap = TRUE) +
  scale_color_manual(values = c("FALSE" = "#2e7d5b", "TRUE" = "#c0392b"),
                     labels = c("Bird-like", "Construction-like")) +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(title = "Timing fingerprint by species",
       subtitle = "Bottom-right (daytime-heavy, weekend-light) behaves like construction",
       x = "Share of detections in weekday business hours",
       y = "Share of detections on weekends",
       color = NULL, size = "Detections") +
  theme_minimal(base_size = 12)
```

```{r construction-hourly-profiles}
# Hour-of-day shape for the flagged species vs everything else. Birds should
# show a dawn bump; construction-like sources ramp up at ~7am and stop by ~6pm.
flagged <- suspect %>% filter(construction_suspect) %>% pull(common_name)

if (length(flagged) > 0) {
  profile <- det %>%
    mutate(group = ifelse(common_name %in% flagged,
                          common_name, "All other species")) %>%
    count(group, hour) %>%
    group_by(group) %>%
    mutate(share = n / sum(n)) %>%   # normalise so shapes are comparable
    ungroup()

  ggplot(profile, aes(hour, share)) +
    annotate("rect", xmin = biz_start, xmax = biz_end, ymin = 0, ymax = Inf,
             alpha = 0.08, fill = "#c0392b") +
    geom_col(fill = "#2e7d5b") +
    facet_wrap(~group, scales = "free_y") +
    scale_x_continuous(breaks = seq(0, 24, 6)) +
    scale_y_continuous(labels = percent) +
    labs(title = "Hourly profile: flagged species vs the rest",
         subtitle = "Shaded band = weekday construction window (7am–6pm)",
         x = "Hour of day", y = "Share of that group's detections") +
    theme_minimal(base_size = 11)
} else {
  cat("No species met the construction-suspect criteria.\n")
}
```

**Reading the result.** A flagged species is a *candidate* for review, not a
verdict — pull its clips and listen (the `clip_url` column links the audio, and
the dashboard's vote buttons let the crowd weigh in). Genuine daytime-active,
city-tolerant birds (e.g. House Sparrow, Rock Pigeon) can also look daytime-heavy,
so weigh the dawn and weekend shares and the median confidence together before
concluding it's a jackhammer.

## 12. How do the species differ? A distributional view of timing

The construction check in §11 was one hand-built rule. Here we step back and ask
the general question: **how does each species distribute itself across the day,
and which ones are unusual?** We build a temporal *fingerprint* per species, then
attack it three ways — Kolmogorov–Smirnov tests (who breaks from the community
rhythm), clustering (how species group), and multivariate outlier scoring
(isolation forest + Mahalanobis).

We use a continuous **fractional hour** (hour + minutes/60) so the distributions
are smooth rather than 24 spikes. Only species with enough detections are
included, since a handful of detections can't describe a distribution.

```{r temporal-keep}
min_n <- 30   # minimum detections for a species to get a fingerprint
keep  <- det %>% count(common_name) %>% filter(n >= min_n) %>% pull(common_name)

det_k <- det %>%
  filter(common_name %in% keep) %>%
  mutate(frac_hour = hour + minute(detected_at) / 60 + second(detected_at) / 3600)

sprintf("%d species clear the n >= %d bar (of %d heard in total).",
        length(keep), min_n, n_distinct(det$common_name))
```

```{r temporal-guard, results="asis"}
# Everything below needs a handful of species to compare. Bail out cleanly (so
# Knit still succeeds) if the data is still too thin.
enough <- length(keep) >= 4
if (!enough) {
  cat("> **Not enough species yet** for a distributional comparison — keep the",
      "listener running and re-knit once more species pass the threshold.\n")
}
```

### 12.1 Per-species temporal fingerprints

Each species becomes a row of features: a circular mean hour and *concentration*
(0 = active all day, 1 = active at one precise time), the entropy of its hourly
profile (high = spread out), and its weekend / dawn / business-hours shares. The
circular sine/cosine components are kept as the model-facing version of "mean
hour" (they avoid the midnight wrap-around that would confuse 23:00 and 00:00).

```{r temporal-fingerprints, eval=enough}
two_pi <- 2 * pi

# 24-bin normalised hour profile per species (each row sums to 1).
H <- det_k %>%
  count(common_name, hour) %>%
  complete(common_name, hour = 0:23, fill = list(n = 0)) %>%
  arrange(common_name, hour) %>%
  group_by(common_name) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  select(common_name, hour, share) %>%
  pivot_wider(names_from = hour, values_from = share)

Hmat <- as.matrix(H[, -1])
rownames(Hmat) <- H$common_name

# Profile entropy, normalised to 0..1 (1 = perfectly flat across 24 hours).
hour_entropy <- apply(Hmat, 1, function(p) {
  p <- p[p > 0]
  -sum(p * log(p)) / log(24)
})

feat <- det_k %>%
  group_by(common_name) %>%
  summarise(
    n             = n(),
    ang_C         = mean(cos(frac_hour / 24 * two_pi)),
    ang_S         = mean(sin(frac_hour / 24 * two_pi)),
    weekend_share = mean(is_weekend),
    dawn_share    = mean(is_dawn),
    business_share = mean(is_business),
    .groups = "drop"
  ) %>%
  mutate(
    concentration = sqrt(ang_C^2 + ang_S^2),
    mean_hour     = (atan2(ang_S, ang_C) %% two_pi) / two_pi * 24,
    hour_entropy  = hour_entropy[common_name]
  )

feat %>%
  transmute(common_name, n, mean_hour, concentration, hour_entropy,
            dawn_share, business_share, weekend_share) %>%
  arrange(mean_hour) %>%
  kable(digits = 3,
        col.names = c("Species", "n", "Mean hour", "Concentration", "Entropy",
                      "Dawn", "Business", "Weekend"),
        caption = "Temporal fingerprint per species (sorted by mean hour of day).")
```

### 12.2 Kolmogorov–Smirnov: who breaks from the community rhythm?

For each species we run a two-sample K-S test of *its* detection times against
the times of **all other species pooled**. A large K-S statistic *D* means its
daily timing departs strongly from the community baseline — exactly the species
worth a closer look (an odd niche, or a non-bird source).

```{r temporal-ks, eval=enough}
all_hours <- det_k$frac_hour

ks_one <- function(sp) {
  x <- det_k$frac_hour[det_k$common_name == sp]
  y <- det_k$frac_hour[det_k$common_name != sp]
  t <- suppressWarnings(ks.test(x, y))   # ties (shared seconds) are harmless here
  data.frame(common_name = sp, D = unname(t$statistic), p_value = t$p.value)
}

ks <- do.call(rbind, lapply(keep, ks_one)) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%   # control false discovery
  arrange(desc(D))

ks %>%
  kable(digits = c(0, 3, 4, 4),
        col.names = c("Species", "K-S D", "p", "p (BH-adj.)"),
        caption = "Departure from the community's daily rhythm (largest first).")

ggplot(head(ks, 15), aes(reorder(common_name, D), D)) +
  geom_col(fill = "#2e7d5b") +
  coord_flip() +
  labs(title = "Most distinctive daily timing (K-S vs all other species)",
       x = NULL, y = "K-S statistic D") +
  theme_minimal(base_size = 12)
```

> Caveat: hour-of-day is *circular*, and K-S treats it as a line — it can
> understate differences that straddle midnight. The concentration metric and the
> clustering below complement it.

### 12.3 Clustering species by daily rhythm

Treat each species' 24-bin hourly profile as a vector and cluster them (Ward's
method on Euclidean distance). The dendrogram shows which species keep the same
hours; the heatmap shows the actual shapes, rows ordered by the clustering.

```{r temporal-cluster, eval=enough, fig.height=6}
d  <- dist(Hmat)                          # distance between hourly profiles
hc <- hclust(d, method = "ward.D2")
k  <- min(4, nrow(Hmat) - 1)              # number of rhythm groups
cl <- cutree(hc, k = k)

plot(hc, hang = -1, main = "Species clustered by daily rhythm",
     xlab = "", sub = "", cex = 0.8)
rect.hclust(hc, k = k, border = "#2e7d5b")
```

```{r temporal-heatmap, eval=enough, fig.height=6}
ord <- rownames(Hmat)[hc$order]
heat_df <- data.frame(common_name = rownames(Hmat), Hmat, check.names = FALSE) %>%
  pivot_longer(-common_name, names_to = "hour", values_to = "share") %>%
  mutate(hour = as.integer(hour),
         common_name = factor(common_name, levels = ord))

ggplot(heat_df, aes(hour, common_name, fill = share)) +
  geom_tile() +
  scale_fill_gradient(low = "#f3f8f5", high = "#1b5e3f", labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "Hourly activity profile by species",
       subtitle = "Rows ordered by the dendrogram above",
       x = "Hour of day", y = NULL, fill = "Share") +
  theme_minimal(base_size = 11)

tibble(common_name = names(cl), rhythm_group = cl) %>%
  arrange(rhythm_group, common_name) %>%
  kable(col.names = c("Species", "Rhythm group"),
        caption = "Cluster membership (groups are data-driven, not labelled).")
```

### 12.4 Outliers: which species' timing is most unusual?

Finally, score each species' full fingerprint for *multivariate* anomaly — not
just one axis, but the whole feature vector at once. **Isolation forest**
(`isotree`) isolates points that are easy to separate from the rest; we also
compute a **Mahalanobis distance** as an always-available baseline. Species
scoring high on both are the genuine temporal oddballs.

> Mahalanobis needs comfortably more species than features to be stable; with
> only a handful (as here) the covariance is singular and it automatically falls
> back to plain (Euclidean) distance from the community centre. Isolation forest
> degrades more gracefully, so it's the score to lean on — but with very few
> species *any* anomaly method is suggestive, not conclusive.

```{r temporal-outliers, eval=enough}
# Model-facing feature matrix (circular components, not the wrap-prone mean hour).
X <- feat %>%
  select(ang_C, ang_S, concentration, hour_entropy,
         dawn_share, business_share, weekend_share)
Xs <- scale(as.matrix(X))
rownames(Xs) <- feat$common_name

# Mahalanobis distance from the community centroid. Falls back to plain distance
# from the centroid if the covariance is singular (highly collinear features).
md <- tryCatch(
  mahalanobis(Xs, colMeans(Xs), cov(Xs)),
  error = function(e) sqrt(rowSums(scale(Xs, scale = FALSE)^2))
)

# Isolation forest, if the optional package is installed.
if (requireNamespace("isotree", quietly = TRUE)) {
  set.seed(1)
  iso <- isotree::isolation.forest(as.data.frame(Xs), ntrees = 200, ndim = 1)
  iso_score <- predict(iso, as.data.frame(Xs))   # 0..1, higher = more anomalous
} else {
  iso_score <- NA_real_
  message("Install the 'isotree' package for isolation-forest scores.")
}

outliers <- feat %>%
  transmute(common_name, n, mean_hour, concentration) %>%
  mutate(mahalanobis = md[common_name],
         iso_score   = if (all(is.na(iso_score))) NA_real_ else iso_score) %>%
  arrange(desc(mahalanobis))

outliers %>%
  kable(digits = 3,
        col.names = c("Species", "n", "Mean hour", "Concentration",
                      "Mahalanobis", "Isolation-forest"),
        caption = "Temporal outliers (most unusual fingerprint first).")
```

```{r temporal-pca, eval=enough}
# A 2-D map of the fingerprint space: nearby species keep similar hours; the
# colour shows how anomalous each one is.
pca <- prcomp(Xs)
pc  <- as.data.frame(pca$x[, 1:2])
pc$common_name <- rownames(Xs)
pc$mahalanobis <- md

ggplot(pc, aes(PC1, PC2, color = mahalanobis)) +
  geom_point(size = 3) +
  geom_text(aes(label = common_name), vjust = -0.8, size = 3,
            check_overlap = TRUE) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Species in temporal-fingerprint space (PCA)",
       subtitle = "Brighter = more unusual daily timing",
       color = "Mahalanobis") +
  theme_minimal(base_size = 12)
```

**How to use this.** The three views agree on the genuinely odd species: a high
K-S *D*, its own little cluster, and a high anomaly score together are a strong
signal that something times its "calls" unlike any real bird in the set — your
prime suspects for construction, machinery, or a misfiring ID. As always, the
clip audio (`clip_url`) is the final arbiter.

## 13. Disconnect

```{r disconnect}
dbDisconnect(con)
```

---

*The Ramble Register — acoustic bird monitoring, Upper West Side, Manhattan.*
*Edit the queries above or add your own chunks; `det` and `votes` hold all the data.*
